$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-05 paper/paper.tex
a pass on contributions

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\todo{highlight that we can talk about a resource even if it has been exhausted (e.g. we can send 0 multiplicities of a channel over another channel}
We live in a concurrent world where any given state is often interactively computed by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, predict, and build such interactive systems; as mathematicians, abstraction is our tool of choice to do so.

The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data over communication channels.
Notably, it features channel mobility: channels themselves are first order values, and can therefore be transmitted.
The \picalc{} has been a foundation for the design and implementation of programming languages for concurrency, such as Pict \cite{Pierce} and (more recently) Go \cite{Golang}.
At the state-of-the-art, the \picalc{} features a wide plethora of types \cite{K07}: basic types, linear types, types for liveness properties (such as deadlock freedom, livelock freedom or termination), and most notably, session types. This makes the \picalc{} a fully-fledged tool for modelling and verification of concurrent and distributed systems.

@@

\subsection{Contributions}
\begin{enumerate}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
 \begin{itemize}
   \item \textbf{Syntax}: \autoref{syntax} uses type level de Bruijn indices \cite{deBruijn1972} to introduce a syntax that is well-scoped by construction: every free variable is accounted for in the type of the process that uses it.

   \item \textbf{Semantics}: \autoref{semantics} provides an operational semantics for the \picalc{}, prior to any typing.
   This operational semantics is given as a reduction relation on processes (in \autoref{operational-semantics}).
   The reduction relation tracks at the type level what channel the communication occurs on, so that this information can then be used to state subject reduction --- aka type preservation (in \autoref{subject-reduction}).
   The reduction relation is defined modulo structural congruence (in \autoref{structural-congruence}) --- a relation that acts as a quotient type that removes unnecessary syntactic minutiae introduced by the syntax of the \picalc{}.
 \end{itemize}

  \item \textbf{Leftover typing for the \picalc{}, with linear, gradual and shared types}:
  \autoref{type-system} provides the type system for a resource-aware \picalc{}.
  \begin{itemize}
    \item \textbf{Algebras on multiplicities}: The type system is \emph{resource-aware} in that it allows the user to provide resource-aware algebras that are then applied to the type-system (in \autoref{multiplicities}).
    Linear types, gradual types and shared types are all instances of such a resource-aware algebra.
    Any \emph{partial commutative monoid} that is \emph{functional}, \emph{cancellative} and has a minimal element is a valid such algebra.
    Multiple algebras can be simultaneously used in a single type system --- usage contexts keep information about what algebra to use on which element (in \autoref{contexts}).

    \item \textbf{Leftover typing}: The typing judgments use \emph{leftover typing} to model the resource-aware \picalc{} (in \autoref{leftover-typing}).
    This approach adds a leftover usage context to the typing judgments.
    Typing derivations take the resources of their input usage context, consume some of them, and leave the rest as leftovers in the output usage context.
    One of the benefits of this technique is that it removes the need for extrinsic context splits, which are rendered unnecessary.
  \end{itemize}
  \item \textbf{Machine verified proofs of subject reduction and auxiliary theorems}:
  \autoref{type-safety} details the type safety properties of our \picalc{} with leftovers.
  Leftover typing allows \emph{framing} (\autoref{thm:framing}), \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}) theorems to be stated.
  We use these to prove \emph{subject congruence} (\autoref{subject-congruence}), and together with \emph{substitution} (\autoref{substitution}), to prove \emph{subject reduction} (\autoref{subject-reduction}).

  \item \textbf{Fully mechanised formalisation available in Agda}:
  This work has been fully formalised in Agda and is publicly available \cite{Zalakain2020Agda}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\todo{highlight that we can talk about a resource even if it has been exhausted (e.g. we can send 0 multiplicities of a channel over another channel}
We live in a concurrent world where any given state is often interactively computed by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, predict, and build such interactive systems; as mathematicians, abstraction is our tool of choice to do so.
\todo{rename gradual to something else...}
The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data over communication channels.
Notably, it features channel mobility: channels themselves are first order values, and can therefore be transmitted.
The \picalc{} has been a foundation for the design and implementation of programming languages for concurrency, such as Pict \cite{Pierce} and (more recently) Go \cite{Golang}.
At the state-of-the-art, the \picalc{} features a wide plethora of types \cite{K07}: basic types, linear types, types for liveness properties (such as deadlock freedom, livelock freedom or termination), and most notably, session types. This makes the \picalc{} a fully-fledged tool for modelling and verification of concurrent and distributed systems.

@@

\subsection{Contributions}
\begin{enumerate}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
 \begin{itemize}
   \item \textbf{Syntax}: \autoref{syntax} uses type level de Bruijn indices \cite{deBruijn1972} to introduce a syntax that is \emph{well-scoped by construction}: every free variable is accounted for in the type of the process that uses it.

   \item \textbf{Semantics}: \autoref{semantics} provides an operational semantics for the \picalc{}, prior to any typing.
   This operational semantics is given as a reduction relation on processes (\autoref{operational-semantics}).
   The reduction relation tracks at the type level the channel the communication occurs on, so that this information can then be used to state subject reduction --- aka type preservation (\autoref{subject-reduction}).
   The reduction relation is defined modulo structural congruence (\autoref{structural-congruence}) --- a relation that acts as a quotient type that removes unnecessary syntactic minutiae introduced by the syntax of the \picalc{}.
 \end{itemize}

  \item \textbf{Leftover typing for the \picalc{} with linear, gradual and shared types}:
  \autoref{type-system} provides the type system for a resource-aware \picalc{}.
  \begin{itemize}
    \item \textbf{Algebras on multiplicities}: The user can provide \emph{resource-aware} algebras, which are then applied to the type system (\autoref{multiplicities}).
    Linear types, gradual types and shared types are all instances of the same resource-aware algebra, giving rise to three different type systems under the same framework.
    Any \emph{partial commutative monoid} that is \emph{functional}, \emph{cancellative} and has a minimal element is a valid such algebra.

    Multiple algebras can be simultaneously used in a single type system --- usage contexts keep information about what algebra to use on which element (\autoref{contexts}).
    This allows for type systems combining linear, gradual and shared types.

    \item \textbf{\picalc{} with leftovers}: Our type system uses \emph{leftover typing} to model the resource-aware \picalc{} (\autoref{leftover-typing}).
    This approach adds a leftover usage context to the typing judgments.
    Typing derivations take the resources of their input usage context, consume some of them, and leave the rest as leftovers in the output usage context.
    One of the benefits of this technique is that it removes the need for extrinsic context splits, which are rendered unnecessary.
  \end{itemize}
  \item \textbf{Machine verified proofs of subject reduction and auxiliary theorems}:
  \autoref{type-safety} details the type safety properties of our \picalc{} with leftovers.
  Leftover typing allows \emph{framing} (\autoref{thm:framing}), \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}) theorems to be stated.
  We use these to prove \emph{subject congruence} (\autoref{subject-congruence}), and together with \emph{substitution} (\autoref{substitution}), to prove \emph{subject reduction} (\autoref{subject-reduction}).

  \item \textbf{Fully mechanised formalisation available in Agda}:
  This work has been fully mechanised in Agda and is publicly available in our GitHub repository \cite{Zalakain2020Agda}.
